---
title             : "Generalizing from Static to Dynamic Face-Orientations"
shorttitle        : "Title"

author: 
  - name          : "Sebastian Montesinos"
    affiliation   : "1"
    corresponding : yes    # Define only one corresponding author
    address       : "221 Peretsman Scully Hall, Princeton University, Princeton, New Jersey"
    email         : "sm6511@princeton.edu"
    role: # Contributorship roles (e.g., CRediT, https://credit.niso.org/)
      - "Conceptualization"
      - "Writing - Original Draft Preparation"
      - "Writing - Review & Editing"
  - name          : "Shruti Japee"
    affiliation   : "2"
    role:
      - "Writing - Review & Editing"
      - "Supervision"
  - name          : "Lina Teichmann"
    affiliation   : "2"
    role:
      - "Writing - Review & Editing"
      - "Supervision"
  - name          : "Chris Baker"
    affiliation   : "2"
    role:
      - "Writing - Review & Editing"
      - "Supervision"

affiliation:
  - id            : "1"
    institution   : "Princeton University"
  - id            : "2"
    institution   : "National Institute of Mental Health"

authornote: |
  Add complete departmental affiliations for each author here. Each new line herein must be indented, like this line.

  Enter author note here.

abstract: |
  One or two sentences providing a **basic introduction** to the field,  comprehensible to a scientist in any discipline.
  Two to three sentences of **more detailed background**, comprehensible  to scientists in related disciplines.
  One sentence clearly stating the **general problem** being addressed by  this particular study.
  One sentence summarizing the main result (with the words "**here we show**" or their equivalent).
  Two or three sentences explaining what the **main result** reveals in direct comparison to what was thought to be the case previously, or how the  main result adds to previous knowledge.
  One or two sentences to put the results into a more **general context**.
  Two or three sentences to provide a **broader perspective**, readily comprehensible to a scientist in any discipline.
  
  <!-- https://tinyurl.com/ybremelq -->
  
keywords          : "keywords"
wordcount         : "X"

bibliography      : "r-references.bib"

floatsintext      : no
linenumbers       : no
draft             : yes
mask              : no

figurelist        : no
tablelist         : no
footnotelist      : no

classoption       : "doc"
output            : papaja::apa6_pdf
---

```{r setup, include = FALSE}
library("papaja")
r_refs("r-references.bib")
```

```{r analysis-preferences}
# Seed for random number generation
set.seed(42)
knitr::opts_chunk$set(cache.extra = knitr::rand_seed)
```



# Methods
We report how we determined our sample size, all data exclusions (if any), all manipulations, and all measures in the study. <!-- 21-word solution (Simmons, Nelson & Simonsohn, 2012; retrieved from http://ssrn.com/abstract=2160588) -->

## Participants

## Material

## Procedure

## Data analysis
We used `r cite_r("r-references.bib")` for all our analyses.

```{r}
# Read in data and plot
data <- read.csv("galton_child_parent_heights.csv")
```

# Results

(ref:my-figure-caption) Shows the relationship between parent and child height using a linear regression. 

```{r my-figure, fig.cap = "(ref:my-figure-caption)"}

library(ggplot2)
ggplot(data, aes(x = child_ht, y = parent_ht)) +
  geom_smooth(method = 'lm') + 
  geom_point() +
  annotate('text', label = "Corr = 0.4587332", y = 74, x = 63) +
  theme_apa()


```


```{r}
  model1 <- lm(child_ht ~ parent_ht, data = data)
```

Statistics for linear model for Figure 1:

`r apa_print(model1)$statistic`


# Discussion


\newpage

# References

::: {#refs custom-style="Bibliography"}
:::
